iSEE: Interactive SummarizedExperiment Explorer

### 2018-04-24 manuscript/main.tex
first draft of ban format, ported entirely to f1000res format

\documentclass{bioinfo}
\copyrightyear{2015} \pubyear{2015}
\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}

\begin{document}
\firstpage{1}

\subtitle{Data and text mining}

\title[iSEE]{iSEE: Interactive SummarizedExperiment Explorer}
\author[Rue-Albrecht \textit{et~al}.]{Kevin Rue-Albrecht\,$^{\text{\sfb 1,}\dagger}$,
    Federico Marini\,$^{\text{\sfb 2,3,}\dagger}$,
    Charlotte Soneson\,$^{\text{\sfb 4,}\dagger}$,
    and Aaron T. L. Lun\,$^{\text{\sfb 5,}\ast,\dagger}$}
\address{$^{\text{\sf 1}}$Kennedy Institute of Rheumatology, University of Oxford, Oxford OX3 7FY, United Kingdom, \\
$^{\text{\sf 2}}$Center for Thrombosis and Hemostasis (CTH), Mainz, \\
$^{\text{\sf 3}}$Institute for Medical Biostatistics, Epidemiology and Informatics, Mainz, \\
$^{\text{\sf 4}}$University of Zurich and SIB Swiss Institute of Bioinformatics, Zurich, Switzerland and \\
$^{\text{\sf 5}}$Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom\\
$^{^\dagger}$These authors contributed equally.}

\corresp{$^\ast$To whom correspondence should be addressed.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{\textbf{Summary:} Data exploration is critical to the comprehension of large biological datasets generated by high-throughput assays such as sequencing.
However, most existing tools for interactive visualization are limited to specific assays or analyses.
Here, we present the iSEE (Interactive SummarizedExperiment Explorer) software package, which provides a general visual interface for exploring data in a SummarizedExperiment object.
iSEE is directly compatible with many existing R/Bioconductor packages for analyzing high-throughput biological data,
and provides useful features such as simultaneous examination of multiple data aspects, dynamic linking between plots and code tracking for reproducibility.
We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics datasets.\\
\textbf{Availability:} iSEE is publicly available as an R package from the open-source Bioconductor project (https://bioconductor.org/packages/iSEE).\\
\textbf{Contact:} \href{aaron.lun@cruk.cam.ac.uk}{aaron.lun@cruk.cam.ac.uk}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

\maketitle

\section{Introduction}
Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays such as transcriptomics.
Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).

@@

\section{Description of the iSEE interface}
Here, we present the iSEE software package for interactive data exploration.
iSEE is implemented in R using the shiny framework \citep{chang2017shiny} and exploits data architectures from the open-source Bioconductor project \citep{gentleman2004bioconductor}.
The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
This object stores one or more matrices of experimental observations as ``assays'', where columns and rows represent samples and features, respectively.
Per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; this includes experimental metadata as well as any analysis results.
All of these data aspects can be directly examined in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
The flexibility of the SummarizedExperiment class is the driving factor behind its broad deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C analysis pipelines \citep{lun2016infrastructure}, amongst others.
By accepting a SummarizedExperiment object, iSEE immediately offers interactive visualization for a variety of data modalities, complementing the state-of-the-art analysis methodologies already available in R/Bioconductor packages.
%In particular, we leverage off the state-of-the-art allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).

@@

\section{Conclusion}
iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics datasets.
We provide a number of demonstrations at https://github.com/LTLA/iSEE2018, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells \citep{bodenmiller2012cytof}.

\section*{Acknowledgements}
We thank the organizers and participants of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
We also thank members of the Bioconductor community for their helpful suggestions.
\vspace*{-12pt}

\section*{Funding}
ATLL was supported by core funding from Cancer Research UK (award no. 17197 to Dr.\ John Marioni).
The work of FM is supported by the German Federal Ministry of Education and Research (BMBF 01EO1003).
\vspace*{-12pt}

%\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
%\bibliographystyle{plain}
%
\bibliography{ref}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Welcome to Overleaf --- just edit your article on the left,
% and we'll compile it for you on the right. If you give 
% someone the link to this page, they can edit at the same
% time. See the help menu above for more info. Enjoy!
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% For more detailed article preparation guidelines, please see:
% http://f1000research.com/author-guidelines

\documentclass[10pt,a4paper,twocolumn]{article}
\usepackage{f1000_styles}

%% Default: numerical citations
\usepackage[numbers]{natbib}

%% Uncomment this lines for superscript citations instead
% \usepackage[super]{natbib}

%% Uncomment these lines for author-year citations instead
% \usepackage[round]{natbib}
% \let\cite\citep

\begin{document}

\title{iSEE: Interactive SummarizedExperiment Explorer}
\titlenote{The title should be detailed enough for someone to know whether the article would be of interest to them, but also concise. Please ensure the broadness and claims within the title are appropriate to the content of the article itself.}
\author[1]{Kevin Rue-Albrecht}
\author[2,3]{Federico Marini}
\author[4]{Charlotte Soneson}
\author[5]{Aaron T. L. Lun}

\affil[1]{Kennedy Institute of Rheumatology, University of Oxford, Oxford OX3 7FY, United Kingdom}
\affil[2]{Center for Thrombosis and Hemostasis (CTH), University Medical Center of the Johannes Gutenberg University Mainz, Mainz}
\affil[3]{Institute for Medical Biostatistics, Epidemiology and Informatics, University Medical Center of the Johannes Gutenberg University Mainz, Mainz}
\affil[4]{University of Zurich and SIB Swiss Institute of Bioinformatics, Zurich, Switzerland}
\affil[5]{Cancer Research UK Cambridge Institute, University of Cambridge, Cambridge CB2 0RE, United Kingdom}

% TODO: $\dagger$ These authors contributed equally.}
% TODO: symbol for corresponding author e.g. $\ast$

\maketitle
\thispagestyle{fancy}

Please list all authors that played a significant role in the research involved in the article. Please provide full affiliation information (including full institutional address, ZIP code and e-mail address) for all authors, and identify who is/are the corresponding author(s).

% TODO: full address and emails too?

\begin{abstract}

\textbf{Summary:} Data exploration is critical to the comprehension of large biological datasets generated by high-throughput assays such as sequencing.
However, most existing tools for interactive visualization are limited to specific assays or analyses.
Here, we present the iSEE (Interactive SummarizedExperiment Explorer) software package, which provides a general visual interface for exploring data in a SummarizedExperiment object.
iSEE is directly compatible with many existing R/Bioconductor packages for analyzing high-throughput biological data,
and provides useful features such as simultaneous examination of multiple data aspects, dynamic linking between plots and code tracking for reproducibility.
We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics datasets.\\
\textbf{Availability:} iSEE is publicly available as an R package from the open-source Bioconductor project (https://bioconductor.org/packages/iSEE).\\
\textbf{Contact:} aaron.lun@cruk.cam.ac.uk\\
\textbf{Supplementary information:} Supplementary data are available online.

% Abstracts should be up to 300 words and provide a succinct summary of the article. Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasize the importance of the work described in the article. Citations should not be used in the abstract, and the use of abbreviations should be minimized. If you are writing a Research or Systematic Review article, please structure your abstract into Background, Methods, Results, and Conclusions.

\end{abstract}

\section*{Keywords}

Please list up to eight keywords to help readers interested in your article find it more easily.

% TODO: pick up nice keywords!

\clearpage

\section*{Introduction}
Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays such as transcriptomics.
Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).

@@

\section*{Description of the iSEE interface}
Here, we present the iSEE software package for interactive data exploration.
iSEE is implemented in R using the shiny framework \citep{chang2017shiny} and exploits data architectures from the open-source Bioconductor project \citep{gentleman2004bioconductor}.
The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
This object stores one or more matrices of experimental observations as ``assays'', where columns and rows represent samples and features, respectively.
Per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; this includes experimental metadata as well as any analysis results.
All of these data aspects can be directly examined in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
The flexibility of the SummarizedExperiment class is the driving factor behind its broad deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C analysis pipelines \citep{lun2016infrastructure}, amongst others.
By accepting a SummarizedExperiment object, iSEE immediately offers interactive visualization for a variety of data modalities, complementing the state-of-the-art analysis methodologies already available in R/Bioconductor packages.
%In particular, we leverage off the state-of-the-art allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).

@@

\section*{Conclusion}
iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics datasets.
We provide a number of demonstrations at https://github.com/LTLA/iSEE2018, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells \citep{bodenmiller2012cytof}.

\subsection*{Author contributions}
In order to give appropriate credit to each author of an article, the individual
contributions of each author to the manuscript should be detailed in this section. We
recommend using author initials and then stating briefly how they contributed.

\subsection*{Competing interests}
All financial, personal, or professional competing interests for any of the authors that
could be construed to unduly influence the content of the article must be disclosed and
will be displayed alongside the article.

\subsection*{Grant information}
ATLL was supported by core funding from Cancer Research UK (award no. 17197 to Dr.\ John Marioni).
The work of FM is supported by the German Federal Ministry of Education and Research (BMBF 01EO1003).

\subsection*{Acknowledgements}
We thank the organizers and participants of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
We also thank members of the Bioconductor community for their helpful suggestions.

{\small\bibliographystyle{unsrtnat}
\bibliography{ref}}

\bigskip
References can be listed in any standard referencing style that uses a numbering system
(i.e. not Harvard referencing style), and should be consistent between references within
a given article.

Reference management systems such as Zotero provide options for exporting bibliographies as Bib\TeX{} files. Bib\TeX{} is a bibliographic tool that is used with \LaTeX{} to help organize the user's references and create a bibliography. This template contains an example of such a file, \texttt{sample.bib}, which can be replaced with your own. Use the \verb|\cite| command  to create in-text citations, \cite{zheng2017massively}
%like this \cite{Smith:2012qr} and this \cite{Smith:2013jd}.

% See this guide for more information on BibTeX:
% http://libguides.mit.edu/content.php?pid=55482&sid=406343

% For more author guidance please see:
% http://f1000research.com/author-guidelines

% When all authors are happy with the paper, use the 
% ‘Submit to F1000Research' button from the menu above
% to submit directly to the open life science journal F1000Research.

% Please note that this template results in a draft pre-submission PDF document.
% Articles will be professionally typeset when accepted for publication.

% We hope you find the F1000Research Overleaf template useful,
% please let us know if you have any feedback using the help menu above.
